The Key to a Data Parallel Compiler

### 2016-04-04 2016-spring-array/2016-array-key.tex
2016 Spring ARRAY submission

\documentclass[pldi]{sigplanconf-pldi15}
% \usepackage[utf8]{inputenc}
\usepackage{array}
\usepackage{fontspec}
\setmainfont[Ligatures=TeX]{STIX}
\setmonofont{APL385 Unicode}
\usepackage[scaled]{helvet} % see www.ctan.org/get/macros/latex/required/psnfss/psnfss2e.pdf
\usepackage{url}                  % format URLs
\usepackage{listings}          % format code
\usepackage{enumitem}      % adjust spacing in enums
\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
% known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink
\newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}   % print a hyperlinked DOI

@@

@@

\begin{itemize}[noitemsep]
\item We describe a method of computing over arbitrary sub-trees selected by their parent-child 
relationships in a data-parallel manner using the Key operator and Node Coordinates.
\item We situate this technique into a broader, language-driven compiler construction strategy 
that enables the development of parallel by construction compilers that exhibit desirable programmability 
features in terms of code size, complexity, portability, and maintainability. 
\item We provide analysis of these techniques as used in a commercial compiler project called Co-dfns and 
report on the results.
\item We compare these techniques to the Nanopass approach to compiler design in terms of programmer 
productivity and compiler runtime performance. 
\item We enumerate a set of programming techniques, patterns, and strategies that complement the 
Key operator and Node Coordinates in the construction of tree transformations in general and give some 
intuitions on the use of these techniques on arbitrary transformations.
\end{itemize}
%[cf]
%[of]:Generalized Sub-tree Computation
\section{Data Parallel Sub-tree Computation}

@@

The only extra issue involved here is to ensure that the order of evaluation matches. In our case, we 
are assuming that the order of evaluation is right to left, which means that the above order is actually 
backwards of the desired order. During recombination, we simply reverse these orders and this fixes 
that problem. More work would be required to take into consideration a specific precedence hierarchy. 
%[cf]
%[of]:Lexical Scoping
\subsection{Lexical Scoping}
%[cf]
%[cf]
%[of]:Table of Compiler Passes
\renewcommand{\arraystretch}{1.5}
\begin{table*}
\centering
\begin{tabular}{l p{2.8in} l l l}
\textbf{Pass} & \textbf{Description} & \textbf{Core Patterns} 
 & \textbf{Coordinate?} \\ \hline
Record Node Coordinates & Adds a new field to each node containing that node's 
 node coordinate. & Outer Product/Scan & Yes \\
Record Function Depths & Adds a new field to each node recording how many 
 functions surround the node. & Inner Product & Yes \\
Drop Unnamed Functions & Eliminates some code that will not be evaluated at 
 the top-level. & Filter & No \\
Drop Unreachable Code & Eliminates some unreachable code. &
 Filter/Inner Product & Yes \\
Lift Functions & Moves all functions to the top-level. & Key & Yes \\
Drop Redundant Nodes & Eliminates unnecessary nodes/nesting & Filter & No \\
Flatten Expressions & Removes nesting from expressions. & Key & Yes \\
Compress Atomic Nodes & Atomizes nested atom nodes & Amend & No \\
Propagate Constants & Inlines all references to literal values. &
 Amend/Rank & Yes \\
Fold Constants & Converts constant expressions to literals & Amend & No \\
Compress Expressions & Converts expression sub-trees into single nodes. & 
 Amend & No \\
Record Final Return Value & Records the value returned by each function. & 
 Key & No \\
Normalize Values Field & Normalizes the shape and size of the values field. &
 Amend & No \\
Lift Type-checking & Infers some type information at compile time &
 Power Limit/Rank & No \\
Allocate Value Slots & Does a form of frame allocation for variables &
 Key & No \\
Anchor Variables & Resolves lexically scoped variables &
 Key & Yes \\
Record Live Variables & Records the variables that are live at each point of 
 execution. & Key & No \\
Fuse Scalar Loops & Identifies Fusion opportunities and fuses expressions &
 Key & No \\
Type Specialization & Specializes each function for a series of potential inputs. 
 & Key & No \\
\end{tabular}
\caption{A listing of some compiler passes in the Co-dfns compiler and their 
 relationship with the Key operator and associated tree computation techniques}
\label{table:passes}
\end{table*}
%[cf]
%[of]:Other Passes
\section{Other Passes}

@@

%[of]:Unintegrated Sections
\section{Unintegrated Sections}

%[of]:Scaling up, Considerations and Thoughts
\subsection{Scaling up, Considerations and Thoughts}

How well do these techniques work in practice? We have applied these techniques to an entire compiler, 
with the same restrictions as used here. The result was a success, and the core of the compiler, which 
is everything but parsing and code generation, is contained in a single file consisting of only 40 lines. 
In 12 point font, the compiler easily fits on a single sheet of paper. It produces C and CUDA code.

Not every pass in the compiler requires this sort of operation. Indeed, if flattening occurs early and 
in the correct order, it’s possible to use much simpler methods of grouping using only a prefix sum 
scan instead of operating over the node coordinates. It’s also interesting to note that we do not 
need to recompute the node coordinates throughout the compiler. Indeed, this would defeat their 
purpose somewhat, as the node coordinates allow us to understand the original position of a node, 
even if that node has now been lifted elsewhere. We use this information when resolving lexical scope, 
for instance.

When writing data-parallel code, duplicated work is common. The existing compiler contains little to 
no inherently duplicated work. Most of the duplicated work exists for brevity and clarity of the code, 
rather than for any required reason, and could easily be factored out.

Finally, it is interesting to note that the simplified control flow gives rise to some positive side-effects. 
Besides brevity and directness, the complexity of each operation and the composition of operations is 
well known. We have not yet tried to do this, but it is straightforward to derive a worst-case complexity 
bound on code written in this style, so simple, in fact, that it can be accomplished automatically. It is 
conceivable then, that the compiler could produce not only an executable program from code written 
in this style, but also a complexity bound on its execution. 
%[cf]
%[of]:Computing with Node Coordinates
\subsection{Computing with Node Coordinates}

Given the above keys and grouping information, we now have all the information we need to do 
function lifting and expression flattening, but we lack a suitable operator to complete the task. 
Indeed, it isn’t immediately apparent that a single, general purpose operator over arrays exists 
which can accomplish such a task. However, the Key operator, while masquerading as a useful business 
analytics operator, represents just such a useful tree transformation tool. 
%[cf]
%[of]:Introduction (Old)
\subsection{Introduction (Old)}

When compiling a high-level language into a low-level language, we often implement a variety of 
flattening transformations to normalize more complex code exhibiting nesting into flat programs 
that are semantically equivalent. Function lifting, for example, takes a program with nested functions 
into an equivalent program in which all functions are at the root of the program. The following function
f always returns 14.

\begin{verbatim}
f←λ.(λ ⍵.(⍵+⍵))7
\end{verbatim}

A function lifting pass would transform this into the following:

\begin{verbatim}
fn1←λ⍵.⍵+⍵
f←λ.fn1 7
\end{verbatim}

Expression flattening works in a similar fasion. Assuming all expressions evaluate right to left, this expression:

\begin{verbatim}
v←(a+b)÷c×d
\end{verbatim}

Becomes this:

\begin{verbatim}
v←c×d
t←a+b
v←t÷v
\end{verbatim}

Characteristic of many compiler passes, most of the work is manipulating the tree, rather than, say, number 
crunching. Almost universally, compilers use a combination of recursion and branching to implement these 
transformations, such as using the visitor pattern, Nanopass framework [19], or direct structural recursion. 
The “work” happens in the control flow of the program text. The ubiquity of this approach warrants note. 
Let us savour such rare solidarity for a moment. Now, let us challenge it.

Let us arbitrarily restrict ourselves to a data-flow, data-parallel model. Specifically, consider a language 
whose core primitives are pointwise functions over n-dimensional arrays. We have a set of operators that 
take functions and apply them over different sub-parts of arrays. We can compose these functions together. 
That is all. Can we implement a compiler in such a language? The obvious answer is, yes, given arbitrary 
primitives. What if we restrict ourselves to standard and general-purpose primitives? The answer is still 
yes. Furthermore, the method is surprisingly direct and concise. The language we use is a subset of Dyalog 
APL 14.0 and the reader may follow the examples in the following sections using TryAPL.org.

The trick to such a compiler is solving the issue illustrated by the previous examples, expressing tree 
manipulation without embedding the core logic in the control flow. We accomplish this by choosing an 
appropriate representation of the AST and a property of each node we call its node coordinate. These 
coordinates allow us to locally answer the inter-node questions we need without complex control flow. When 
we combine this with a general-purpose operator called “Key,” we can perform arbitrary computation over 
sub-trees that we partition based on these inter-node relationships, such as whether one node is an ancestor 
of another. When combined with more mundane array programming, we can implement the complete core of a 
compiler, including lexical scope, function lifting, expression flattening/normalization, loop fusion, 
frame/register allocation, and so on. The result is a completely data-parallel compiler modulo parsing and 
some aspects of code generation.

The notation used throughout this paper is introduced piecemeal as necessary in each relevant section in 
which the notation is first used. All functions are applied infix and take either one or two arguments. All 
expressions evaluate from right to left. Tables 1 and 2 provide a reference for the notation and primitives 
as a reference aid.

%[of]:Contributions
\paragraph{Contributions}

\begin{itemize}[noitemsep]
\item We introduce node coordinates as a means to locally compute inter-node properties without using complex 
control flow.
\item We demonstrate methods to perform arbitrary computation over sub-trees partitioned by these 
inter-node properties in a pure data-flow, data-parallel style.
\item The presented methods are direct and concise; the data structures, simple.
\item These methods require no specialized operations, but instead, use only established, general-purpose 
array programming.
\end{itemize}
%[cf]
%[cf]

%[cf]

>>>>

\documentclass[numbers,preprint]{sigplanconf}
% \usepackage[utf8]{inputenc}
\usepackage{array}
\usepackage{fontspec}
\setmainfont[Ligatures=TeX]{STIX}
\setmonofont{APL385 Unicode}
\usepackage[scaled]{helvet} % see www.ctan.org/get/macros/latex/required/psnfss/psnfss2e.pdf
\usepackage{url}                  % format URLs
\usepackage{listings}          % format code
\usepackage{enumitem}      % adjust spacing in enums
\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
% known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink
\newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}   % print a hyperlinked DOI
\usepackage{booktabs}

@@

\conferenceinfo{ARRAY '16}{June, 2016, Santa Barbara, CA, USA}
\copyrightyear{2016}

\authorinfo{Aaron W. Hsu}{Indiana University}{awhsu@indiana.edu}

@@

\begin{itemize}[noitemsep]
\item We describe a method of computing over arbitrary sub-trees selected by their parent-child 
relationships in a data-parallel manner using the Key operator and Node Coordinates.
\item We situate this technique into a broader, language-driven compiler construction strategy 
that enables the development of parallel by construction compilers that exhibit desirable programmability 
features in terms of code size, complexity, portability, and maintainability. 
\item We provide analysis of these techniques as used in a commercial compiler project called Co-dfns and 
report on the results.
\item We compare these techniques to the Nanopass approach to compiler design in terms of programmer 
productivity and compiler runtime performance. 
\item We enumerate a set of programming techniques, patterns, and strategies that complement the 
Key operator and Node Coordinates in the construction of tree transformations in general and give some 
intuitions on the use of these techniques on arbitrary transformations.
\end{itemize}

\begin{table}
\centering
\begin{tabular}{cll}
\toprule
Symbol                   & Monadic            & Dyadic \\
\midrule
\texttt{+}               & Identity           & Plus (Add) \\
\texttt{-}               & Negative           & Minus (Subtract) \\
\texttt{×}               & Direction (Signum) & Times (Multiply) \\
\texttt{÷}               & Reciprocal         & Divide \\
\texttt{|}               & Magnitude          & Residue (Modulo) \\
\texttt{⌊}               & Floor              & Minimum \\
\texttt{⌈}               & Ceiling            & Maximum \\
\texttt{*}               & Exponential        & Power \\
\texttt{⍟}               & Natural Logarithm  & Logarithm \\
\texttt{○}               & Pi Times           & Circular (Trigonometric) \\
\texttt{!}               & Factorial          & Binomial \\
\texttt{\textasciitilde} & Not                & \\
\texttt{?}               & Roll               & \\
\texttt{\^}              &                    & And \\
\texttt{∨}               &                    & Or \\
\texttt{⍲}               &                    & Nand \\
\texttt{⍱}               &                    & Nor \\
\texttt{<}               &                    & Less \\
\texttt{≤}               &                    & Less Or Equal \\
\texttt{=}               &                    & Equal \\
\texttt{≥}               &                    & Greater Or Equal \\
\texttt{>}               &                    & Greater \\
\texttt{≠}               &                    & Not Equal \\
\end{tabular}
\caption{Scalar Primitives in Proposed Language}
\label{tab:scalarprims}
\end{table}

\begin{table}
\centering
\begin{tabular}{cll}
\toprule
Symbol     & Monadic       & Dyadic \\
\midrule
\texttt{⍴} & Reshape       & \\
\texttt{,} & Ravel         & Catenate/Laminate \\
\texttt{⍪} & Table         & Catenate First/Laminate \\
\texttt{⌽} & Reverse       & Rotate \\
\texttt{⊖} & Reverse First & Rotate First \\
\texttt{⍉} & Transpose     & Transpose \\
\texttt{↑} & Mix           & \\
\texttt{↓} & Split         & \\
\texttt{⊂} & Enclose       & Partitioned Enclose \\
\texttt{∊} & Enlist        & \\
\end{tabular}
\caption{Structural Mixed Primitives in Proposed Language}
\label{tab:structprims}
\end{table}

\begin{table}
\centering
\begin{tabular}{cll}
\toprule
Symbol                   & Monadic  & Dyadic \\
\midrule
\texttt{⊃}               & Disclose & Pick\\
\texttt{↑}               &          & Take \\
\texttt{↓}               &          & Drop \\
\texttt{/}               &          & Replicate \\
\texttt{⌿}               &          & Replicate First \\
\texttt{\textbackslash}  &          & Expand \\
\texttt{⍀}               &          & Expand First \\
\texttt{\textasciitilde} &          & Without (Excluding) \\
\texttt{∩}               &          & Intersection \\
\texttt{∪}               & Unique   & Union \\
\texttt{⊣}               & Same     & Left \\
\texttt{⊢}               & Identity & Right\\
\end{tabular}
\caption{Selection Mixed Primitives in Proposed Language}
\label{tab:selprims}
\end{table}

\begin{table}
\centering
\begin{tabular}{cll}
\toprule
Symbol     & Monadic         & Dyadic \\
\midrule
\texttt{⍳} & Index Generator & Index Of \\
\texttt{∊} &                 & Membership \\
\texttt{⍋} & Grade Up        & Grade Up \\
\texttt{⍒} & Grade Down      & Grade Down \\
\texttt{?} &                 & Deal \\
\texttt{⍷} &                 & Find \\
\end{tabular}
\caption{Selector Mixed Primitives in Proposed Language}
\label{tab:seltorprims}
\end{table}

\begin{table}
\centering
\begin{tabular}{cll}
\toprule
Symbol     & Monadic       & Dyadic \\
\midrule
\texttt{⍴} & Shape         & \\
\texttt{≡} & Depth         & Match \\
\texttt{≢} & Tally         & Not Match \\
\texttt{⍎} & Execute       & Execute \\
\texttt{⍕} & Format        & Format \\
\texttt{⊥} &               & Decode (Base) \\
\texttt{⊤} &               & Encode (Representation) \\
\texttt{⌹} & Matrix Divide & Matrix Inverse \\
\end{tabular}
\caption{Miscellaneous Mixed Primitives in Proposed Language}
\label{tab:miscprims}
\end{table}

\begin{table*}
\centering
\begin{tabular}{cll}
\toprule
Symbol     & Name & Description \\
\midrule
\texttt{⍨} & Commute & Swaps arguments or distributes
 right argument to both sides \\
\texttt{¨} & Each & Applies its operand point-wise over the left/right
 arguments \\
\texttt{/} & Reduce & Reduce along the last axis \\
\texttt{⌿} & Reduce First & Reduce along the first axis \\
\texttt{\textbackslash} & Scan & Scan along the last axis \\
\texttt{⍀} & Scan First & Scan along the first axis \\
\texttt{⌸} & Key & Apply operand once for each sub-array grouped by key \\
\end{tabular}
\caption{Primitive Adverbs in Proposed Language, each takes a single left
operand and describes a verb operating over one or two arguments}
\label{tab:adverbs}
\end{table*}

\begin{table*}
\centering
\begin{tabular}{cll}
\toprule
Symbol     & Name & Description \\
\midrule
\texttt{∘} & Compose & Composes two operands as in traditional mathematics \\
\texttt{.} & Inner Product & Inner product operation, e.g.
 \texttt{+.×} for matrix multiplication \\
\texttt{∘.} & Outer Product & Cartesian product or ``function table'' \\
\texttt{⍣} & Power & Iteration, Limited use only \\
\texttt{⍤} & Rank & Apply a verb along cells of an array \\
\end{tabular}
\caption{Primitive Conjunctions in Proposed Language, each takes a left
 and right operand and describes a verb operating over one or two arguments}
\label{tab:conjunctions}
\end{table*}

\section{Data Parallel Sub-tree Computation}

@@

The only extra issue involved here is to ensure that the order of evaluation matches. In our case, we 
are assuming that the order of evaluation is right to left, which means that the above order is actually 
backwards of the desired order. During recombination, we simply reverse these orders and this fixes 
that problem. More work would be required to take into consideration a specific precedence hierarchy. 
%[cf]
%[of]:Lexical Scoping
\subsection{Lexical Scoping}
%[cf]
%[cf]
%[of]:Table of Compiler Passes
\renewcommand{\arraystretch}{1.5}
\begin{table*}
\centering
\begin{tabular}{l p{2.8in} l l l}
\toprule
Pass & Description & Core Patterns & Coordinate? \\
\midrule
Record Node Coordinates & Adds a new field to each node containing that node's 
 node coordinate. & Outer Product/Scan & Yes \\
Record Function Depths & Adds a new field to each node recording how many 
 functions surround the node. & Inner Product & Yes \\
Drop Unnamed Functions & Eliminates some code that will not be evaluated at 
 the top-level. & Filter & No \\
Drop Unreachable Code & Eliminates some unreachable code. &
 Filter/Inner Product & Yes \\
Lift Functions & Moves all functions to the top-level. & Key & Yes \\
Drop Redundant Nodes & Eliminates unnecessary nodes/nesting & Filter & No \\
Flatten Expressions & Removes nesting from expressions. & Key & Yes \\
Compress Atomic Nodes & Atomizes nested atom nodes & Amend & No \\
Propagate Constants & Inlines all references to literal values. &
 Amend/Rank & Yes \\
Fold Constants & Converts constant expressions to literals & Amend & No \\
Compress Expressions & Converts expression sub-trees into single nodes. & 
 Amend & No \\
Record Final Return Value & Records the value returned by each function. & 
 Key & No \\
Normalize Values Field & Normalizes the shape and size of the values field. &
 Amend & No \\
Lift Type-checking & Infers some type information at compile time &
 Power Limit/Rank & No \\
Allocate Value Slots & Does a form of frame allocation for variables &
 Key & No \\
Anchor Variables & Resolves lexically scoped variables &
 Key & Yes \\
Record Live Variables & Records the variables that are live at each point of 
 execution. & Key & No \\
Fuse Scalar Loops & Identifies Fusion opportunities and fuses expressions &
 Key & No \\
Type Specialization & Specializes each function for a series of potential inputs. 
 & Key & No \\
\end{tabular}
\caption{A listing of some compiler passes in the Co-dfns compiler and their 
 relationship with the Key operator and associated tree computation techniques}
\label{table:passes}
\end{table*}
%[cf]
%[of]:Other Passes
\section{Other Passes}

@@
